\chapter*{Abstract}

\paragraph{What is this thesis about?}
This thesis will present a comprehensive study on the role of visual perceptual segmentation related to gaze prediction and its dynamics. It aims to explore how segmentation affects visual perception and lay the foundation for future research on neural dynamics in visual processing in the context of perceptual segmentation.
We will investigate the interplay between segmentation and visual perception through a series of experiments and analyses, ultimately contributing to a deeper understanding of the underlying neural mechanisms.

\paragraph{Why is this thesis important?}

The importance of this thesis lies in its potential to advance our understanding of visual perception and segmentation. By examining the neural dynamics associated with these processes, we can gain insights into how the brain interprets and organizes visual information.\\
This research has implications for various fields, including cognitive neuroscience, psychology, and artificial intelligence, where understanding visual perception is crucial for developing better models and applications.\\
Most importantly this research lays solid foundations for a running hypothesis that visual perceptual segmentation is a key factor in gaze prediction and neural dynamics, and its findings will be part of a larger body of work that aims to unravel the complexities of visual processing in the human brain.

\paragraph{What will be the structure of the thesis}
We will begin with an introduction to the topic, outlining the significance of visual perceptual segmentation and its relevance to gaze prediction. The thesis will then delve into the methodology used in our research, including experimental design, data collection, and analysis techniques.\\

\paragraph{Results}

Thanks to the works done in this thesis, we will present a new State of the Art (SoTA) model for gaze prediction that outperforms existing methods. This model will be based on the findings of our research, demonstrating the significance of visual perceptual segmentation in gaze prediction tasks.\\

\thispagestyle{empty}
\mbox{}
\newpage
